---
title: "Case study #2"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
pacman::p_load(magrittr, dplyr, tidyverse, ggplot2, table1, VGAM, brant, MASS)
ggthemr::ggthemr("fresh")
```

For the second case study you will be predicting self-rated health category for individuals in their midlife (`outcome` variable). Select the correct model based on the distribution of the outcome variable. In step 1 you will conduct the main effects model with the variables `depress`, `alcage`, `cigage`, `age`, and `bp` predicting the variable `health`. In step 2 you will report the predicted probabilities for each self-rated health category among those with diagnosed high or low blood pressure, respectively. In step 3 you will test any relevant assumptions. Based on the results of Step 3, what would you recommend as the next step in predicting self-rated health category?

```{r}
case2 <- rio::import("https://github.com/mp5782/RegressionFinal/blob/main/MIDUS%20III%20Final%20Exam%20Fall%202021.dta?raw=true")

#clean the data to exclude "don't know", "refused", "inapp" responses from the alcage, ciage variables
case2 %<>% mutate_at(c("alcage", "cigage"), function(x) ifelse(x>96, NA, x))

# make the outcome variable as ordered factor (increasing order as health improves)
case2 %<>% mutate(health = ordered(health, levels = c(4, 3, 2, 1)))

# to account for those who never drank or smoke, create a corresponding dichotomous variable
case2 %<>% mutate(
  smoke = ifelse(cigage==96, 0, 1),
  drink = ifelse(alcage==96, 0, 1)
)

```

# Step 1. Univariate analysis

```{r}
#outcome variable
ggplot(case2, aes(health))+geom_bar()

# explanatory variables
table1(~factor(depress) + factor(bp) + factor(smoke) + factor(drink) + age + alcage + cigage, data = case2)
```

# Step 2. Bivariate analysis

```{r}
# table
table1(~factor(depress) + factor(bp) + age + alcage + cigage | health, data = case2)

# univaraite cumulative logit
for(var in c("depress", "bp", "age", "alcage", "cigage")){
  temp <- case2[which(case2[,var]!=96),]
  print(
    vglm(case2$health ~ case2[,var], family=cumulative(parallel=TRUE)) %>% summary()
  )
}
```

# Step 3. Fit the main effect regression model

```{r}
main_mod  <- vglm(health ~ depress + alcage + cigage + age +bp, 
                        family=cumulative(parallel=TRUE), data=case2)

summary(main_mod)

temp<-summary(main_mod)
# Exponentiated coefficients
res <- data.frame(
coefficients = round(temp@coefficients,3),
# 95% CI
ci_low = round(temp@coef3[,1]-1.96*temp@coef3[,2],3),
ci_high = round(temp@coef3[,1]+1.96*temp@coef3[,2], 3),
#p-value
pval =round(temp@coef3[,4],3)
)

res
exp(res)
```

# Step 4. Proportional odds assumption test

```{r}
# check for proportional odds assumption
polr_mod <- polr(health ~ depress + alcage + cigage + age +bp,
                 Hess = T, data = case2)
brant(polr_mod)
```

# Step 5. Model fit (vs. the null model)

```{r}
lrtest(main_mod)
```

# Step 5. Predicted probability

```{r}
fitted_bp <- data.frame(bp = case2$bp, fitted(main_mod))
pb <- fitted_bp %>% group_by(bp) %>% summarize(mean(X1), mean(X2), mean(X3), mean(X4))

pb
pb %>% as.data.frame() %>% gather("health", "p", -bp) %>% mutate(
  health = case_when(health=="mean(X1)" ~ "Excellent",
                     health=="mean(X2)" ~ "Very good/Good",
                     health=="mean(X3)" ~ "Fair",
                     TRUE ~ "Poor"),
  bp = ifelse(bp==1, "High blood pressure", "No high blood pressure")
) %>% ggplot(aes(health, p)) + geom_bar(aes(group = bp, fill = factor(bp)),stat="identity", position = "dodge") +
  theme(legend.title = element_blank(),
        legend.position = "top")
```
