---
title: "Case Study #1"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
pacman::p_load(magrittr, dplyr, tidyverse, ggplot2, table1, VGAM, brant, MASS,haven,ResourceSelection,knitr)
ggthemr::ggthemr("fresh")
case1 <- read_dta("MIDUS III Final Exam Fall 2021.dta")
```

For the first study you will be predicting if people in their midlife have ever experienced heart trouble (outcome variable). Select the correct model based on the distribution of the outcome variable. In step 1 you will conduct a model with the variables `bp`, `smoke`, `age`, `male`, and `exercise` predicting the variable `heart.` In step 2 you will run the model with the interaction term between `smoke` and `male`, controlling for the other variables. In step 3 you will report at least 3 types of model fit statistics for each individual model, and assess which model (the main effects or interaction effect model) is better. Please use two tests to evaluate if there are significant outliers that may be influential in the model with the better fit.

# Step 0. Univariate + bivariate analysis

```{r 1.0,echo=F,comment=NA}
# all variables
table1(~age+factor(male)+factor(heart)+cigage+factor(smoke)+alcage+factor(depress)+factor(bp)+factor(exercise)+factor(health), data = case1)
#outcome variable 1
ggplot(case1, aes(heart))+geom_bar()
#outcome variable 2
ggplot(case1, aes(health))+geom_bar()
#bivar, case 1
table1(~age+factor(male)+cigage+factor(smoke)+alcage+factor(depress)+factor(bp)+factor(exercise)+factor(health)|factor(heart), data = case1)
#bivar, case 2
table1(~age+factor(male)+factor(heart)+cigage+factor(smoke)+alcage+factor(depress)+factor(bp)+factor(exercise)|factor(health), data = case1)
```

# Step 1. model variables `bp`, `smoke`, `age`, `male`, and `exercise` predicting the variable `heart`.

```{r 1.1, echo=F,comment=NA}
mod1.1 <- glm(heart ~ bp+factor(smoke)+age+factor(male)+factor(exercise), family = binomial, data = case1)
summary(mod1.1)
```

# Step 2. model with the interaction term between `smoke` and `male`, controlling for the other variables.

```{r 1.2, echo=F,comment=NA}
mod1.2 <- glm(heart ~ bp+factor(smoke)+age+factor(male)+factor(exercise)+I(smoke*male), family = binomial, data = case1)
summary(mod1.2)
```

# Step 3. report at least 3 types of model fit statistics for each individual model, and assess which model (the main effects or interaction effect model) is better

ARE WE TALKING ABOUT ONLY TWO DIFFERENT MODELS HERE, OR 5 + 1 + 1 = 7?!

```{r 1.3,echo=F,comment=NA,results='asis'}
#Likelihood Ratio tests for each model vs. null:
a1 <- pchisq(mod1.1$deviance,mod1.1$df.residual)
a2 <- pchisq(mod1.2$deviance,mod1.2$df.residual)
#Hosmer-Lemeshow Goodness of Fit:
b1 <- hoslem.test(mod1.1$y,mod1.1$fitted.values,g=10)
b2 <- hoslem.test(mod1.2$y,mod1.2$fitted.values,g=10)
#AIC of each test
c1 <- AIC(mod1.1)
c2 <- AIC(mod1.2)
#BIC of each test
d1 <- BIC(mod1.1)
d2 <- BIC(mod1.2)
Evaluation <- c("Likelihood Ratio vs. Null","Hosmer&LemeshowGoodness of Fit","AIC","BIC")
mod1 <- c(a1,b1$p.value,c1,d1)
mod2 <- c(a2,b2$p.value,c2,d2)
results <- data.frame("Evaluation"=Evaluation,"Main Effects Model"=mod1,"Interaction Effect Model"=mod2)
kable(results)
```

# Step 4. use two tests to evaluate if there are significant outliers that may be influential in the model with the better fit.

```{r 1.4,echo=F}
case1_0 <- case1[case1$heart==0,] #data, no heart trouble
case1_1 <- case1[case1$heart==1,] #data, heart trouble
#pearson residuals
stdres <- residuals(mod1.2,type='pearson')
stdres_0 <- stdres[case1$heart==0] #pearson residuals, no heart trouble
stdres_1 <- stdres[case1$heart==1] #pearson residuals, heart trouble
plot(case1_0$M2ID,stdres_0,
             main='Pearson Residuals, no heart trouble',
             ylab='standardized Pearson residuals',
             xlab='ID')

plot(case1_1$M2ID,stdres_1,
             main='Pearson Residuals, heart trouble', 
             ylab='standardized Pearson residuals',
             xlab='ID')
#deviance reciduals
deviance.res<-residuals(mod1.2,type='deviance')
deviance.res_0<-deviance.res[case1$heart==0] #deviance residuals, no trouble
deviance.res_1<-deviance.res[case1$heart==1] #deviance residuals, heart trouble
plot(case1_0$M2ID,deviance.res_0,
     ylab='Deviance Residuals',
     xlab='ID')
plot(case1_1$M2ID,deviance.res_1,
     ylab='Deviance Residuals',
     xlab='ID')
```